Top-down versus bottom-up cohesiveness in microbial community coalescence

### 2021-05-25 draft/draft_v3.tex
coalescence.py now shows more plots; more draft written

To better understand the underlying mechanisms that govern
the emergence of ecological co-selection,
we developed a minimal model of community coalescence.
This model is comprised of two communities with two species each
as illustrated in \figref[a]{fig4}.
Within each community, the dominant species
($\mathrm{s}_1$ and $\mathrm{s}_3$ in the resident and invasive communities
respectively)
is able to utilize the primary resource that is supplied externally ($\mathrm{R}_1$).
As a result of $\mathrm{R}_1$ metabolism,
the dominants secrete a single byproduct off which
the sub-dominant species ($\mathrm{s}_2$ and $\mathrm{s}_4$ respectively) can feed.
Finally, these sub-dominants secrete an additional resource that can again
be metabolized by the respective dominants.
This is the simplest structure where both top-dwn and bottom-up
cohesiveness can emerge, with their relative strengths being determined
by how effectively the species can utilize each other's metabolic byproducts.
Several parameters can modulate the direction of community cohesiveness,
and even in this simple model there can be complex interactions between them.
For instance, a sub-dominant that is very efficient at metabolizing the dominant
secretions will therefore be a strong producer of secondary secretions,
in turn increasing the metabolic flow towards the dominant itself.

@@

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.7,keepaspectratio]{figs/fig1.pdf}
\caption{\textbf{Overview of the experimental protocol.}
\textbf{a.}~Environmental samples collected from eight different locations were used
to inoculate our communities.
\textbf{b.}~Communities were stabilized in serial batch culture bioreactors
\cite{Goldford2018} in minimal synthetic media with glutamine or citrate as the
only supplied carbon source.
\textbf{c.}~Communities were plated in minimal media agar plates and the most abundant
species (the ``dominants") from each community were isolated. We refer to the set of
sub-dominant species as the ``cohorts".
\textbf{d.}~Rank-frequency distributions of all eight communities stabilized in either
glutamine (red) or citrate (blue), sequenced at a depth of $10^{-4}$ reads.
Three biological replicates per community are shown.
Community compositions are skewed and long-tailed.
\textbf{e.}~Our hypothesis is that ecological co-selection can take place from the top-down,
i.e. the dominant co-selecting the cohort, or from the bottom-up, i.e. the cohort co-selecting
the dominant. Both forms of co-selection can be positive (recruitment) or negative
(antagonism).
\textbf{f.}~Illustration of the protocol of our coalescence experiments. All pairs of
communities were inoculated into fresh minimal media supplemented with the same carbon
source where communities had been previously stabilized. The coalesced (C) and original
resident (R) and invasive (I) communities were then serially diluted and allowed to grow
for seven additional transfers.}
\label{fig1}
\end{figure}

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.7,keepaspectratio]{figs/fig3.pdf}
\caption{\textbf{Trade offs between bottom-up and top-down ecological co-selection.}
\textbf{a.} We hypothesize that three scenarios are possible regarding bottom-up co-selection:
sub-dominant species could co-select for (green) or against (red) their dominant in coalescence,
or they could have no effect in the invasion success of the dominant taxa (gray).
\textbf{b.} Simulations with a Microbial Consumer-Resource Model suggest that positive bottom-up
co-selection is common and can be very strong, whereas negative bottom-up co-selection is rare.
We plot the frequency reached by the invasive dominants when invading the resident communities
in isolation versus the same frequency when invading together with their cohorts,
i.e. in community coalescence.
Points in the green/red area represent instances where the invasive dominant is able to invade
with higher/lower success when accompanied by its cohort, evidencing positive/negative
bottom-up co-selection.
Points around the diagonal (gray area) correspond to cases where the success of the invasive dominant
is only weakly affected by the presence or absence of its cohort.
\textbf{c.} We divided the data from our simulations into two sets according to whether
positive or no bottom-up co-selection was observed (that is, whether points fell into the green or gray
areas of panel b). Here we reproduce the plots in \figref[b]{fig2} for each set, representing the
result of the dominant head-to-head pairwise competition versus the outcome of community coalescence.
Left panel: when positive bottom-up co-selection is strong, head-to-head pairwise competition of dominants
is poorly predictive of coalescence outcomes
($R^2=0.00$, $p>0.05$).
Right panel: on the other hand, when bottom-up co-selection is weak coalescence outcomes are more
strongly dictated by the result of the dominant-dominant competition
($R^2=0.34$, $p<10^{-6}$).
\textbf{d.} Experiments show that in our conditions, positive bottom-up co-selection is indeed more
frequent and strong than negative bottom-up co-selection.
\textbf{e.} We reproduce the plots in panel c for our experimental data. This is equivalent to recreating
\figref[b]{fig2}, this time splitting our data by the strength of bottom-up co-selection
instead of by the carbon source provided to the communities.
Like simulations predicted, instances of community coalescence displaying strong positive bottom-up co-selection
yield a poor correspondence between dominant pairwise competition and coalescence outcomes
(left panel, $R^2=0.07$, $p>0.05$)
and vice-versa
(right panel, $R^2=0.37$, $p<10^{-4}$).}
\label{fig3}
\end{figure}

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.7,keepaspectratio]{figs/fig4_v2.pdf}
\caption{\textbf{A minimal model of community coalescence.}
\textbf{a.} Blablabla...
\textbf{b.} Blablabla...}
\label{fig4}
\end{figure}

>>>>

To better understand the underlying mechanisms that govern
the emergence of ecological co-selection,
we developed a minimal model of community coalescence.
This model is comprised of two communities with two species each
as illustrated in \figref[a]{fig4}.
Within each community, the dominant species
($\mathrm{s}_1$ and $\mathrm{s}_3$ in the resident and invasive communities
respectively)
is able to utilize the primary resource that is supplied externally ($\mathrm{R}_1$).
As a result of $\mathrm{R}_1$ metabolism,
the dominants secrete a single byproduct
($\mathrm{R}_2$ and $\mathrm{R}_4$ respectively)
off which
the sub-dominants ($\mathrm{s}_2$ and $\mathrm{s}_4$ respectively) can feed.
Finally, these sub-dominants secrete an additional resource
($\mathrm{R}_3$ and $\mathrm{R}_5$ respectively)
that can again
be metabolized by the respective dominants.
This is the simplest structure where both top-down and bottom-up
cohesiveness can emerge, with their relative strengths being determined
by how effectively the species can utilize each other's metabolic byproducts.
Several parameters can modulate the strength and direction of community cohesiveness,
and even in this simple model there can be complex interactions between them.
For instance, a sub-dominant that is very efficient at metabolizing the dominant's
secretions will therefore be a strong producer of secondary byproducts,
in turn increasing the metabolic flow towards the dominant itself.
To keep both communities symmetrical and
mantain tractability of the model's behavior, we chose to make
the two sub-dominant species equally able consumers of the
byproduct of their respective dominants
(\methodsref{sim-min}).
We also made it so species secrete only one type of byproduct each,
i.e. the metabolic matrix is binary in this case
(see \hyperref[box1]{Box~1}).
The secretions of the resident and invasive communities are
non-overlapping:
\iffalse
(i.e. $\mathrm{s}_1$ and $\mathrm{s}_2$ secretions
differ from $\mathrm{s}_3$ and $\mathrm{s}_4$ secretions)
\fi
this ensures that coalescence always results in one of the two
communities taking over and completely excluding the other one,
which facilitates the interpretation of the model's outcomes.

With these constraints, the model is specified by four rates:
the resident dominant's uptake rate of the primary resource ($c_{11}$)
and of the secondary byproduct secreted by $\mathrm{s}_2$ ($c_{13}$),
as well as
the invasive dominant's uptake rate of the primary resource ($c_{31}$)
and of the secondary byproduct secreted by $\mathrm{s}_4$ ($c_{35}$).
Analogous to what we did for our initial
experiments and simulations of community coalescence,
we first stabilized each community separately
through serial growth-dilution cycles,
and then mixed them 1:1 and stabilized again until generational
equilibrium was achieved.
Equations \ref{eq:crm-n} and \ref{eq:crm-r} in \hyperref[box1]{Box~1}
describe the dynamics of this system between consecutive dilutions
with the corresponding choices for the parameter values
(\methodsref{sim-min}).
We ran simulations for both community-community competitions and
invasions from the invasive dominant alone.
We considered four different scenarios illustrated in \figref[b-e]{fig4}:
\textit{1)} both communities are cohesive
from the top-down but not from the bottom-up,
\textit{2)} the resident community is cohesive
both from the top-down and from the bottom-up
while
the resident community is cohesive
from the top-down only,
\textit{3)} the resident community is cohesive
from the top-down only
while
the invasive community is cohesive
both from the top-down and from the bottom-up,
and \textit{4)} both communities are cohesive
from the top-down and from the bottom-up.
The outcomes in each case are discussed below.

\paragraph*{Top-down resident versus top-down invasive.}
This is the simplest scenario.
When both communities are cohesive strictly from the top-down
($c_{13}=0$ and $c_{35}=0$),
coalescence outcomes are determined by which dominant is a better consumer
of the primary resource.
The most competitive dominant excludes the least competitive one
($\mathrm{s}_1$ excludes $\mathrm{s}_3$ if $c_{11}>c_{31}$ and vice-versa,
\figref[b]{fig4})
and,
because the cross-feeding networks of the invasive and resident
communities are non-overlapping,
it co-selects its sub-dominant exclusively.
The relative similarity between the invasive and the coalesced communities
($Q$, see \methodsref{metrics})
can only be 0 (if the resident dominant ``wins'' competition for
$\mathrm{R}_1$) or 1 (if the opposite happens).
The invasion success of the invasive dominant ($\mathrm{s}_3$) is
unaffected by the presence or absence of its sub-dominant ($\mathrm{s}_4$)
since it is not being cross-fed by it.
Thus, again $\mathrm{s}_3$ is only able to invade alone when it
outcompetes the resident dominant ($\mathrm{s}_1$) for the primary resource ($c_{11}>c_{31}$).

\paragraph*{Bidirectional resident versus top-down invasive.}
Blablabla...

\paragraph*{Top-down resident versus bidirectional invasive.}
Blablabla...

\paragraph*{Bidirectional resident versus bidirectional invasive.}
Blablabla...

@@

\subsubsection*{Minimal model}\label{methods:sim-min}

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.7,keepaspectratio]{figs/fig1.pdf}
\caption{\textbf{Overview of the experimental protocol.}
\textbf{a.}~Environmental samples collected from eight different locations were used
to inoculate our communities.
\textbf{b.}~Communities were stabilized in serial batch culture bioreactors
in minimal synthetic media with glutamine or citrate as the
only supplied carbon source.
\textbf{c.}~Communities were plated in minimal media agar plates and the most abundant
species (the ``dominants") from each community were isolated. We refer to the set of
sub-dominant species as the ``cohorts".
\textbf{d.}~Rank-frequency distributions of the eight communities stabilized in either
glutamine (red) or citrate (blue), sequenced at a depth of $10^{-4}$ reads.
Three biological replicates per community are shown.
Community compositions are skewed and long-tailed.
\textbf{e.}~Our hypothesis is that ecological co-selection can take place from the top-down,
i.e. the dominant co-selecting the cohort, or from the bottom-up, i.e. the cohort co-selecting
the dominant. Both forms of co-selection can be positive (recruitment) or negative
(antagonism).
\textbf{f.}~Illustration of the protocol of our coalescence experiments. All pairs of
communities were inoculated into fresh minimal media supplemented with the same carbon
source where communities had been previously stabilized. The coalesced (C) and original
resident (R) and invasive (I) communities were then serially diluted and allowed to grow
for seven additional transfers.}
\label{fig1}
\end{figure}

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.7,keepaspectratio]{figs/fig3.pdf}
\caption{\textbf{Trade offs between bottom-up and top-down ecological co-selection.}
\textbf{a.} We hypothesize that three scenarios are possible regarding bottom-up co-selection:
sub-dominant species could co-select for (green) or against (red) their dominant in coalescence,
or they could have no effect in the invasion success of the dominant taxa (gray).
\textbf{b.}
\iffalse
Simulations with a Microbial Consumer-Resource Model suggest that positive bottom-up
co-selection is common and can be very strong, whereas negative bottom-up co-selection is rare.
\fi
Simulations with a Microbial Consumer-Resource Model:
we plot the frequency reached by the invasive dominants when invading the resident communities
in isolation versus the same frequency when invading together with their cohorts,
i.e. in community coalescence.
Points in the green/red area represent instances where the invasive dominant is able to invade
with higher/lower success when accompanied by its cohort, evidencing positive/negative
bottom-up co-selection.
Points around the diagonal (gray area) correspond to cases where the success of the invasive dominant
is only weakly affected by the presence or absence of its cohort.
\textbf{c.} We divided the data from our simulations into two sets according to whether
positive or no bottom-up co-selection was observed (that is, whether points fell into the green or gray
areas of panel b). Here we reproduce the plots in \figref[b]{fig2} for each set, representing the
result of the dominant head-to-head pairwise competition versus the outcome of community coalescence.
Left panel: strong positive bottom-up co-selection
($R^2=0.00$, $p>0.05$).
Right panel: no bottom-up co-selection
($R^2=0.34$, $p<10^{-6}$).
\iffalse
Left panel: when positive bottom-up co-selection is strong, head-to-head pairwise competition of dominants
is poorly predictive of coalescence outcomes
($R^2=0.00$, $p>0.05$).
Right panel: on the other hand, when bottom-up co-selection is weak coalescence outcomes are more
strongly dictated by the result of the dominant-dominant competition
($R^2=0.34$, $p<10^{-6}$).
\fi
\textbf{d.} Experiments show that in our conditions, positive bottom-up co-selection is indeed more
frequent and strong than negative bottom-up co-selection.
\textbf{e.} We reproduce the plots in panel c for our experimental data,
i.e. we recreate
\figref[b]{fig2} but this time splitting our data by the strength of bottom-up co-selection
instead of by the carbon source supplied to the communities.
Left panel: strong positive bottom-up co-selection
($R^2=0.07$, $p>0.05$).
Right panel: no bottom-up co-selection
($R^2=0.37$, $p<10^{-4}$).}
\label{fig3}
\end{figure}

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.7,keepaspectratio]{figs/fig4_v2.pdf}
\caption{\textbf{A minimal model of community coalescence.}
\textbf{a.} Illustration of the model structure and parameters.
The primary resource ($\mathrm{R}_1$) is replenished after
each growth-dilution cycle (red arrows).
Solid arrows indicate resource consumption, dashed arrows represent
resource secretion.
\textbf{b-e.} Coalescence outcomes in the minimal model
under different relations of cohesiveness between the
resident and the invasive communities.}
\label{fig4}
\end{figure}
